The Evolution of Phenotypically Invariant Genetic Networks

### 2017-10-03 kalman_walk_paper.tex
edited envelope calc and kalman

%Any linear and minimal gene network 
%-- minimal, informally meaning that the system's phenotype is achieved with the fewest possible genes ($n=\min$, see \ref{apx:kalman})
%-- has an identical impulse response and thus phenotype given identical inputs $u(t)$, up to a change of coordinates. 
%  \begin{equation}
%    \begin{aligned}
%      h(t) &= C e^{A t} B 
%      = C V^{-1} V e^{A t} V^{-1} V B \\
%      &= C V^{-1} e^{V A V^{-1} t} V B 
%      = \bar{C} e^{\bar{A} t} \bar{B}
%    \end{aligned}
%  \end{equation}
%Two gene networks, $\mathcal{S} = ( A, B, C )$, and $\bar{\mathcal{S}} = (\bar{A} = VAV^{-1}, \bar{B} = VB, \bar{C} = CV^{-1} )$, 
%have the same phenotype, for all possible inputs, if and only if they are related by a change of coordinates. 

@@

%Therefore varying $V$ can tweak the relationships between genes within a regulatory network, yet preserve the phenotype. Some gene networks, however, can grown or shrink, perhaps following gene duplications and deletions, and also still preserve their phenotypes. These changes cannot be described by $\mathcal{S}(V)$, but are ecanpuslated by \ref{eqn:equivalence}, below.

@@

\begin{definition}[Phenotypic equivalence of systems]
    Let $(\kappa(t),\phi(t))$ and $(\bar \kappa(t),\bar \phi(t))$ be the solutions to \eqref{eqn:system}
    with coefficient matrices $(A,B,C)$ and $(\bar A,\bar B,\bar C)$ respectively,
    and both $\kappa(0)$ and $\bar \kappa(0)$ are zero. 
    The systems defined by $(A,B,C)$ and $(\bar A,\bar B,\bar C)$ are
    \textbf{phenotypically equivalent} 
    if
    \begin{align*}
        \phi(t) = \bar \phi(t) \qquad \text{for all} \; t \ge 0.
    \end{align*}
    Equivalently, this occurs if and only if
    \begin{align*}
        h(t) = \bar h(t)  \qquad \text{for all} \; t \ge 0,
    \end{align*}
    where $h$ and $\bar h$ are the impulse responses of the two systems.
\end{definition}

One way to find other systems equivalent to a given one
is by change of coordinates:
if $V$ is an invertible matrix, then the systems $(A,B,C)$ and $(VAV^{-1},VB,CV^{-1})$
have the same dynamics because their impulse responses are equal:
  \begin{equation}
    \begin{aligned}
      h(t) &= C e^{A t} B 
      = C V^{-1} V e^{A t} V^{-1} V B \\
      &= C V^{-1} e^{V A V^{-1} t} V B 
      = \bar{C} e^{\bar{A} t} \bar{B}
    \end{aligned}
  \end{equation}
%\begin{align*}
%    CV^{-1}( zI - VAV^{-1})^{-1}VB
%    =
%    CV^{-1}V( zI - A)^{-1}V^{-1}VB
%    =
%    C( zI - A)^{-1}B .
%\end{align*}
However, the converse is not necessarily true: 
systems can have identical impulse responses without being changes of coordinates of each other.
In fact, systems with identical impulse responses can involve interactions between different
numbers of molecular species, and thus be in different dimensions.

>>>>

Any linear and minimal gene network 
-- minimal, informally meaning that the system's phenotype is achieved with the fewest possible genes ($n=\min$, see \ref{apx:kalman})
-- has an identical impulse response and thus phenotype given identical inputs $u(t)$, up to a change of coordinates. 
  \begin{equation}
    \begin{aligned}
      h(t) &= C e^{A t} B 
      = C V^{-1} V e^{A t} V^{-1} V B \\
      &= C V^{-1} e^{V A V^{-1} t} V B 
      = \bar{C} e^{\bar{A} t} \bar{B}
    \end{aligned}
  \end{equation}
Two gene networks, $\mathcal{S} = ( A, B, C )$, and $\bar{\mathcal{S}} = (\bar{A} = VAV^{-1}, \bar{B} = VB, \bar{C} = CV^{-1} )$, 
have the same phenotype, for all possible inputs, if and only if they are related by a change of coordinates. 

@@

Therefore varying $V$ can tweak the relationships between genes within a regulatory network, yet preserve the phenotype. Some gene networks, however, can grown or shrink, perhaps following gene duplications and deletions, and also still preserve their phenotypes. These changes cannot be described by $\mathcal{S}(V)$.

@@

%\begin{definition}[Phenotypic equivalence of systems]
 %   Let $(\kappa(t),\phi(t))$ and $(\bar \kappa(t),\bar \phi(t))$ be the solutions to \eqref{eqn:system}
  %  with coefficient matrices $(A,B,C)$ and $(\bar A,\bar B,\bar C)$ respectively,
   % and both $\kappa(0)$ and $\bar \kappa(0)$ are zero. 
   % The systems defined by $(A,B,C)$ and $(\bar A,\bar B,\bar C)$ are
   % \textbf{phenotypically equivalent} 
   % if
   % \begin{align*}
   %     \phi(t) = \bar \phi(t) \qquad \text{for all} \; t \ge 0.
   % \end{align*}
   % Equivalently, this occurs if and only if
   % \begin{align*}
   %     h(t) = \bar h(t)  \qquad \text{for all} \; t \ge 0,
   % \end{align*}
   % where $h$ and $\bar h$ are the impulse responses of the two systems.
%\end{definition}

%One way to find other systems equivalent to a given one
%is by change of coordinates:
%if $V$ is an invertible matrix, then the systems $(A,B,C)$ and $(VAV^{-1},VB,CV^{-1})$
%have the same dynamics because their impulse responses are equal:
%  \begin{equation}
%    \begin{aligned}
%      h(t) &= C e^{A t} B 
%      = C V^{-1} V e^{A t} V^{-1} V B \\
%      &= C V^{-1} e^{V A V^{-1} t} V B 
%      = \bar{C} e^{\bar{A} t} \bar{B}
%    \end{aligned}
%  \end{equation}
%\begin{align*}
%    CV^{-1}( zI - VAV^{-1})^{-1}VB
%    =
%    CV^{-1}V( zI - A)^{-1}V^{-1}VB
%    =
%    C( zI - A)^{-1}B .
%\end{align*}
%However, the converse is not necessarily true: 
%systems can have identical impulse responses without being changes of coordinates of each other.
%In fact, systems with identical impulse responses can involve interactions between different
%numbers of molecular species, and thus be in different dimensions.
